## Tapestri Package {#tapestri_library}
The purpose of this section is to demonstrate how to extract the data.  We will use the tapestri package.

The 'tapestri' package is available from Mission Bio [here](https://support.missionbio.com/hc/en-us/articles/360042455574-tapestri-R-Package).  We'll use this to read in our sample files.  You'll notice in a lot of the code, we are going to run loops and lapply over all of the samples. For reproducibility purposes, I am going to show a limited example on 5 samples for preprocessing, and later we will load an rds object with all the data in the paper. 

Critical files used for analysis can be found on google drive [here](https://drive.google.com/open?id=17Zw6Ixu93UM7M5Vyl_aOJ7aX2iYIb8If). We're working on getting all of the processsed and raw data up on dbGAP now. Next, make a project folder and set the working directory to that folder:

```{r set working directory, eval=FALSE}
setwd("/Users/bowmanr/Projects/scDNA")
```

Load in the relevant packages we will use later.
```{r packages ,message=FALSE, warning=FALSE}
options(stringsAsFactors = FALSE)
library(VariantAnnotation)
library(plyranges)
library(BSgenome.Hsapiens.UCSC.hg19)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(dplyr)
library(tidyr)
library(purrr)
library(tapestri)
```

Place the downloaded files into a new folder called "data" also make an "analysis" folder. We'll save things there frequently just to provide checkpoints so everything does not need to be run from scratch.  

```{r eval=FALSE}
system("mkdir /Users/bowmanr/Projects/scDNA/data")
system("mkdir /Users/bowmanr/Projects/scDNA/analysis")
```

The following took <20 minutes for the 6 samples on my mac book, but obivously would take much longer for the whole cohort. The final step using the "convert_to_analyte" function in unnecessary, but you will see it comes into play when we merge with the DNA+protein data later in the paper.

```{r load in NGT files, message=FALSE,eval=FALSE,cache = FALSE}
sample_set <- list.files("./data/",full.names = TRUE)
names(sample_set) <-list.files("./data/")

for(i in names(sample_set)){
  barcode_files<-grep("barcode",list.files(sample_set[i],full.names=TRUE),value=TRUE)
  loom_files<-grep("loom$",list.files(sample_set[i],full.names=TRUE),value=TRUE)
  header_files<-grep("vcf_header.txt$",list.files(sample_set[i],full.names=TRUE),value=TRUE)
  barcodes <- read_barcodes(barcode_files,header_files)
  loom <- connect_to_loom(loom_files)
  ngt_file <- extract_genotypes(loom, barcodes, 
                              gt.filter=TRUE, gt.gqc = 30,
                              gt.dpc = 10, gt.afc = 20,  gt.mv = 50, 
                              gt.mc = 50, gt.mm = 1, gt.mask = TRUE)
  snv <- convert_to_analyte(data=as.data.frame(ngt_file),
                             type='snv',
                             name=i)
  saveRDS(snv,paste0("./analysis/",i,".rds"))
}
```
